---
title: "Test divergence between chains"
output: 
  pdf_document:
    toc: true
---

# Compare across chains

```{r}
library(blme)
library(party)
library(sjPlot)
library(gplots)
library(dplyr)


setwd("/Library/WebServer/Documents/ILMTurk/stats/R")
# The similarity between physical signals across chains
load("stabilityAcrossChains.Rdat")


dx = stability.accross.chains[stability.accross.chains$run!="Nij" ,]
dx = dx[complete.cases(dx),]

dx$distance.norm = dx$distance - mean(dx$distance)
dx$curvature.norm = dx$curvature - 0.2
dx$curvature.norm.q = (dx$curvature.norm + 0.1)^2 
dx$curvature.norm.c = (dx$curvature.norm + 0.1)^3 
dx$gen = dx$gen - 5

summary(lm(distance~curvature*gen, data=dx))
```



```{r}
plotmeans(distance~gen,dx[dx$curvature==0,], main='Motor distance')
plotmeans(distance~gen,dx[dx$curvature==0.1,], add=T, col=2, main='Motor distance')

plotmeans(distance~curvature,dx, main='Motor distance')

plotmeans(acousticDistance~gen,dx[dx$curvature==0,], main='Signal distance')
plotmeans(acousticDistance~gen,dx[dx$curvature==0.5,], add=T, col=2, main='Signal distance')
plotmeans(acousticDistance~curvature,dx, main='Signal distance')
```



```{r}
#dx = dx[dx$curvature<0.4,]

m0 = lmer(distance.norm~ 1 + (1|chainA) + (1|chainB),data=dx)
m1 = lmer(distance.norm~ 1 + curvature.norm + (1|chainA) + (1|chainB),data=dx)
m2 = lmer(distance.norm~ 1 + curvature.norm + gen  + (1|chainA) + (1|chainB),data=dx)
m3 = lmer(distance.norm~ 1 + curvature.norm*gen + (1|chainA) + (1|chainB),data=dx)
m4 = lmer(distance.norm~ 1 + curvature.norm* gen  +curvature.norm.q + (1|chainA) + (1|chainB),data=dx)
m5 = lmer(distance.norm~ 1 + curvature.norm* gen  +curvature.norm.q*gen + (1|chainA) + (1|chainB),data=dx)
m6 = lmer(distance.norm~ 1 + curvature.norm* gen  +curvature.norm.q*gen + curvature.norm.c + (1|chainA) + (1|chainB),data=dx)
m7 = lmer(distance.norm~ 1 + curvature.norm* gen  +curvature.norm.q*gen +curvature.norm.c*gen + (1|chainA) + (1|chainB),data=dx)

anova(m0,m1,m2,m3,m4,m5,m6,m7)
sjp.lmer(m7, 'pred', c("gen","curvature.norm"), facet.grid = F)

px = data.frame(
  curvature.norm = 
    rep(
      unique(dx$curvature.norm), 
      each=length(unique(dx$gen))),
  gen = 
    rep(
      unique(dx$gen),
      length(unique(dx$curvature.norm))))
px$curvature =px$curvature.norm +0.2
px$curvature.norm.q = (px$curvature.norm+ 0.1)^2
px$curvature.norm.c = (px$curvature.norm+ 0.1)^3
px$distance.norm = predict(m5,px, re.form=NA)

xyplot(distance.norm~gen, groups=curvature, px, type='b')

plotmeans(distance.norm~gen, dx[dx$curvature==0.4,])

```


```{r}
dx = stability.accross.chains[stability.accross.chains$run=="MT" ,]
stree = ctree(distance~gen+curvature,data=dx)
plot(stree)
```

\newpage

# Compare within chains

Distance between a signal and its direct ancestor.

```{r}

load("stability_withinChains.Rdat")

stability = stability[stability$run!="Nij",]
stability$chain2 = paste(stability$run,stability$chain)
stability$participant = paste(stability$run, stability$chain, stability$gen)



plotmeans(dtw.dist~gen, stability[stability$curvature==0,])
plotmeans(dtw.dist~gen, stability[stability$curvature==0.5,],add=T,col=2)
plotmeans(dtw.dist~curvature, stability)

sx <- group_by(stability, participant)
stability2 <- summarise(sx,
  gen = mean(gen),
  curvature = mean(curvature),
  chain2 = head(chain2,1),
  dtw.dist = mean(dtw.dist, na.rm = TRUE),
  dtw.dist.acoustic = mean(dtw.dist.acoustic, na.rm = TRUE))

plotmeans(dtw.dist~gen, stability2[stability2$curvature==0,])
plotmeans(dtw.dist~gen, stability2[stability2$curvature==0.5,],add=T,col=2)
plotmeans(dtw.dist~curvature, stability2)

plotmeans(dtw.dist.acoustic~gen, stability2[stability2$curvature==0,])
plotmeans(dtw.dist.acoustic~gen, stability2[stability2$curvature==0.5,],add=T,col=2)
plotmeans(dtw.dist.acoustic~curvature, stability2)

```


```{r}
# Physical signal
dx = stability2
m0 = lmer(dtw.dist~1 + (1 | chain2), data=dx)
m1 = lmer(dtw.dist~curvature + (1 | chain2), data=dx)
m2 = lmer(dtw.dist~curvature + gen + (1 | chain2), data=dx)
m3= lmer(dtw.dist~curvature*gen + (1 | chain2), data=dx)
anova(m0,m1,m2,m3)
```

```{r}
# Acoustic signal
dx = stability2
m0A = lmer(dtw.dist.acoustic~1 + (1 | chain2), data=dx)
m1A = lmer(dtw.dist.acoustic~curvature + (1 | chain2), data=dx)
m2A = lmer(dtw.dist.acoustic~curvature + gen + (1 | chain2), data=dx)
m3A= lmer(dtw.dist.acoustic~curvature*gen + (1 | chain2), data=dx)
anova(m0A,m1A,m2A,m3A)

```



# Other measures

## Signal set distinctiveness

How similar are the signals produced by each participant? e.g. How distinct from each other are the signals that participant 12 produced for meanings 1,2,3?  This is measured using the method of splitting the signal time/pitch space into grids and counting differences.

```{r}
plotmeans(signals_MT$hzSigDistinctiveness~signals_MT$gen,
          ylab='Acoustic signal distinctiveness')
plotmeans(signals_MT$physSigDistinctiveness~signals_MT$gen,
          ylab='Physical signal distinctiveness')

plotmeans(hzSigDistinctiveness~curvature, signals_MT,
          ylab='Acoustic signal distinctiveness')

plotmeans(physSigDistinctiveness~curvature, signals_MT,
          ylab='Physical signal distinctiveness')

plotmeans(hzSigDistinctiveness~gen, signals_MT[signals_MT$curvature==0,],
          ylab='Acoustic signal distinctiveness')

plotmeans(hzSigDistinctiveness~gen, signals_MT[signals_MT$curvature==0.2,],
          ylab='Acoustic signal distinctiveness')

plotmeans(hzSigDistinctiveness~gen, signals_MT[signals_MT$curvature==0.5,],
          ylab='Acoustic signal distinctiveness')


```

